plyranges: a grammar of graphics for genomics data

### 2017-11-27 paper.Rmd
first pass at introduction

# Introduction
## What do we mean by genomic data?

  - How does that relate to 'tidy' data?
  - Why tidy data is a useful abstraction.
  - Why _Ranges_ follow this pattern.

Key references: @Wickham2014-jc, @Lawrence2013-wg

## How can we analyse genomic data?

  - the Ranges data structures
  - other abstractions for analysing genomic data
    - the query language approaches
    - the bedtools approach
    - relational algebras

Key references: @Quinlan2010-gc, @Dale2011-js, @Kaitoua2017-pw, @Ovaska2013-gd, @Mungall2014-dg @Kozanitis2016-bm

## What does plyranges contribute?

`plyranges` API provides

- a consistent grammar for analysing genomic data.
- fluent interface enables readable pipelines 
- reproducibility via code clarity, declaraitve programming,  and return consistency

>>>>

# Introduction

Genomic data may be naturally represented as sets of pairs of integers 
corresponding to the start and end points of sequences. Further information
such as strandedness and chromosome name may be added to these sets to 
provide biological context. Because of the flexibility of this representation 
supplemental information such as measurements obtained from an experimental 
assay or annotations from a genome database can be joined to their relevant
sequences. In the Bioconductor/R packages `IRanges` and `GenomicRanges` these representations have been implemented as a suite of data structures called _Ranges_ (@Lawrence2013-wg). These data structures cover many common data types encountered in bioinformatics analyses - a gene can be represented with its coordinates, along with its identifier and the identifiers of its exons; or
an RNA-seq experiment may be represented as sets of genes with a matching count column.

The Bioconductor infrastructure for computing with genomic ranges are highly 
effecient and powerful, however the application programming interface (API) for performing analysis tasks with _Ranges_ is complex due to its large number of
methods and classes. It also makes resulting scripts written 
difficult for a non-programmer to read and reason about. An alternative approach 
would be to implement a domain specific language (DSL) as a fluent interface 
built on top _Ranges_. The goal of fluent interface is to enable users to write 
human-readable code via method chaining and consistent function returns. Fluent 
interfaces fit naturally in the context of Bioinformatics workflows because they 
enable writing succinct pipelines.

Several other DSLs have been implemented to reason about genomics data. Broadly, 
these are either implemented as query languages or as command line tools 
embedded in the unix environment ^[other ideas to mention GROK @Ovaska2013-gd]. An example of the former is the Genome Query Language (GQL) and its distributed 
implementation GenAp which use an SQL-like syntax for fast retrieval of 
information from genomic databases and BAM files (@Kozanitis2014-va, @Kozanitis2016-bm). 
Another  example is the Genometric Query Language (GMQL) which implements a relational algebra for combining big genomic datasets (@Kaitoua2017-pw).  The command line application 
BEDtools develops an extensive algebra for performing arithmetic between two or 
more sets of genomic regions  (@Quinlan2010-gc). It also has a python interface 
which simplifies construcing scripts for performing analyses based on BEDTools  (@Dale2011-js). ^[probably should also mention something about their cons, and more detaiil]

The abstraction provided by the _Ranges_ data structures aligns with the concept
of tidy data developed by @Wickham2014-jc. The tidy data pattern is useful because it allows us to see how the data relates to the design of an experiment
and the variables measured. Consequently, it makes both the modelling and manipulation of data more systematic. The _Ranges_ data structure follows this abstraction: it is a rectanglular table corresponding to a single biological context. Each row contains a single observation and each column a variable about that observation.

The tidy data abstraction has motivated the development of `plyranges`
a grammar of genomic data manipulation based on the _Ranges_ data structures. It
implements and extends the grammar defined by the R package `dplyr` (@Wickham2017-dplyr). The grammar provides a consistent way of interacting with and analysing genomic data via methods for constructing, grouping, mutating, filtering, and summarising _Ranges_ and an algebra for reasoning about actions on _Ranges_ and relationships between _Ranges_.
